---
title: "Lab 3 OM 566"
author: "Christine Raj"
format: html
editor: visual
---

## Libraries and Packages

```{r}
library(data.table)
library(tidyverse)
library(leaflet)
library(lubridate)
```

## Download the data

```{r}
download.file(
  "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
  destfile = file.path("~", "Downloads", "met_all.gz"),
  method   = "libcurl",
  timeout  = 60
)

met <- data.table::fread(file.path("~", "Downloads", "met_all.gz"))
```

## Checking Dimensions

```{r}
dim(met)
head(met)
tail(met)
```

## Looking at variables

```{r}
str(met)
```

## Key Variables

```{r}
table(met$year)
table(met$day)
table(met$hour)
summary(met$temp)
summary(met$elev)
summary(met$wind.sp)
met[met$elev==9999.0] <- NA
summary(met$elev)
met <- met[temp> -40]
met2 <- met[order(temp)]
head(met2)
```

The highest weather station is at 4113.

## Checking against external sources

```{r}
met <- met[temp > -15]
met2 <- met[order(temp)]
head(met2)
```

## Calculating Summary Statistics

```{r}
elev <- met[elev==max(elev), ]
summary(elev)
cor(elev$temp, elev$wind.sp, use="complete")
cor(elev$temp, elev$hour, use = "complete")
cor(elev$wind.sp, elev$day, use = "complete")
cor(elev$wind.sp, elev$hour, use = "complete")
cor(elev$temp, elev$day, use = "complete")
```

## Exploratory Graphs

```{r}
hist(met$elev, breaks = 100)
hist(met$temp)
hist(met$wind.sp)
met_stations <- unique(met[,c("lat", "lon")])
dim(met_stations)
leaflet(elev) %>% addProviderTiles('OpenStreetMap') %>% addCircles(lat=~lat, lng=~lon, opacity=1, fillOpacity = 1, radius = 100)
elev$date <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$date)
elev <- elev[order(date)]
head(elev)
plot(elev$date, elev$temp, type='l')
plot(elev$date, elev$wind.sp, type='l')
```

The graphs show that in August the temperature at each elevation is fairly similar for highs and lows of each day. For wind speed there are around 2 weeks in the middle of august where wind speed became much higher than the first and last week.

## Other Questions

Is the wind direction similar at the same time each day?

```{r}
hist(met$wind.dir)
elev$hour <- with(elev, ymd_h(paste(year, month, day, hour, sep= ' ')))
summary(elev$hour)
elev <- elev[order(hour)]
plot(elev$hour, elev$wind.dir, type = 'l')
```
