---
title: "Lab4"
author: "Christine Raj"
format: html
editor: visual
---

## Downloading Data

```{r}
if (!file.exists("met_all.gz"))
  download.file(
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/02_met/met_all.gz",
    destfile = "met_all.gz",
    method   = "libcurl",
    timeout  = 60
    )
met <- data.table::fread("met_all.gz")

library(tidyverse)
library(data.table)
library(dplyr)
library(ggplot2)
library(leaflet)
library(gganimate)
```

## Preparing the Data

1.  <div>

    ```{r}
    met <- met[temp>-17]
    met2 <- met[order(temp)]
    head(met2)
    summary(met$elev)
    met[met$elev==9999.0] <- NA
    summary(met$elev)
    met$date <- as.Date(paste(met$year, met$month, met$day, sep = "-"))

    metweek1 <- met[, Week := week(met$date)]
    metweek1 <- filter(metweek1, Week == 31)

    met_avg <- metweek1[,.(
      temp     = mean(temp,na.rm=TRUE),
      rh       = mean(rh,na.rm=TRUE),
      wind.sp  = mean(wind.sp,na.rm=TRUE),
      dew.point  = mean(dew.point),
      vis.dist = mean(vis.dist,na.rm=TRUE),
      lat      = mean(lat),
      lon      = mean(lon), 
      elev     = mean(elev,na.rm=TRUE)
    ), by=c("USAFID", "day")]

    met_avg$region <- ifelse(met_avg$lon < -98.00 & met_avg$lat >= 39.71, "NW",
                               ifelse(met_avg$lon < -98.00 & met_avg$lat <39.71, "SW",
                                      ifelse(met_avg$lon >= -98.00 & met_avg$lat >= 39.71, "NE",
                                             ifelse(met_avg$lon >= -98.00 & met_avg$lat < 39.71, "SE", NA))))
    met_avg[, elev_cat := fifelse(elev > 252, "high", "low")]
    ```

    </div>

## Violin Plot of Wind Speed and Dew Point by Region

```{r}
met_avg[!is.na(region)] %>% 
  ggplot() + 
  geom_violin(mapping = aes(x = 1, y = met_avg$dew.point, color = met_avg$region)) + facet_wrap(~ region, nrow=1)
met_avg[!is.na(region)] %>% 
  ggplot() + 
  geom_violin(mapping = aes(x = 1, y = met_avg$wind.sp, color = met_avg$region)) + facet_wrap(~ region, nrow=1)
```

There is heavier distribution in dew point in the western regions.

There is a wide distribution with potential outliers in the northeast region.

## Jitter and Stat Smooth

```{r}
met_avg[!is.na(region)] %>% 
  ggplot() + 
  geom_jitter(mapping = aes(x = met_avg$wind.sp, y = met_avg$dew.point, color = met_avg$region)) + stat_smooth(mapping = aes(x = wind.sp, y = dew.point))
```

As wind speed gets higher the dew point tends to be slightly lower.

## Barplots of weather stations

```{r}
met_avg[!is.na(region)] %>%
  ggplot() + 
  geom_bar(mapping = aes(x = met_avg$elev_cat, fill = met_avg$region), position = "dodge") + 
  scale_fill_brewer(palette = 4) +
  labs(title = "Weather Station Elevation by Region") + 
  labs(x = expression("Elevation Categories"), y = "Number of Observations at Elevation")
```

The northeast has similar amounts of observations at high and low elevations but the northwest has very few observations at low elevations. The southeast has a very high number of observations at the low elevations.

## Stat Summary

```{r}
met_avg[!is.na(region)] %>%
  ggplot() + 
    stat_summary(mapping = aes(x = met_avg$wind.sp, y = met_avg$region), geom = "errorbar", 
    fun.data = "mean_sdl")
met_avg[!is.na(region)] %>%
  ggplot() + 
    stat_summary(mapping = aes(x = met_avg$dew.point, y = met_avg$region), geom = "errorbar",
    fun.data = "mean_sdl")
```

Dew point average is lower in western regions and higher in eastern regions.

Average wind speed is lower in the eastern regions and higher in the western regions.

## Spatial Trend in relative humidity

```{r}
met_avg2 <- met_avg[,.(rh = mean(rh,na.rm=TRUE), lat = mean(lat), lon = mean(lon)),  by=c("USAFID")]
met_avg2 <- met_avg2[!is.na(rh)]
rh.pal <- colorNumeric(c('turquoise','purple','coral'), domain=met_avg2$rh)
rh.pal

humiditymap <- leaflet(met_avg2) %>% 
  # The looks of the Map
  addProviderTiles('CartoDB.Positron') %>% 
  # Some circles
  addCircles(
    lat = ~met_avg$lat, lng=~met_avg$lon,
    label = ~paste0(round(met_avg2$rh,2)), color = ~ rh.pal(met_avg2$rh),
    opacity = 1, fillOpacity = 1, radius = 500
    ) %>%
  # legend
  addLegend('bottomleft', pal=rh.pal, values=met_avg2$rh,
          title='Humidity', opacity=1)
humiditymap
```

## Ggplot extension

```{r}

ggplot(met_avg, aes(factor(met_avg$dew.point), met_avg$region)) + 
  geom_boxplot() +
  # Here comes the gganimate code
  transition_states(
    gear,
    transition_length = 2,
    state_length = 1
  ) +
  enter_fade() + 
  exit_shrink() +
  ease_aes('sine-in-out')
```
